**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-08-13 paper/manuscript.Rmd
first full draft

---
bibliography: references.bib
link-citations: true
csl: msystems.csl
output:
  pdf_document:
    keep_tex: true
  word_document: default
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{multirow}
---

@@

## Introduction (\~250 words)

Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States [@siegel2020]. Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e. colonoscopy), many people do not comply with recommended screening guidelines [@garcía2011a] prompting a need for low cost, non-invasive detection methods. A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e. advanced adenomas) and carcinomas. Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}. However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown. Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs [@callahan2017]. However, whether the additional resolution provided by ASVs is useful for ML classification is unclear {pat preprint?}. Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. Topçuoğlu *et al* [@topçuolu2020] recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml). This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.

## Results (\~700 words)

@@

Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quantified how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or carcinoma) using five machine learning methods with the Mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU (p \< 0.01), but borderline equivalent to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels. These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.

While AUROC values between models is a useful way to compare the overall model performance across all thresholds, they can be misleading [@lobo2008]. Depending on the intended implementation of the model, one may want to optimize the true-positive rate(or sensitivity) over the false-postivite rate (or specificity). 
To further compare the model performance across taxonomic levels we compared the sensitivity of the models

-   on average, how many positives are missed with ASV?

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels. The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 3). While the resolution provided by ASVs is useful in

|   **TODO: what are important taxa, patterns along levels (e.g fuso always in top )**

Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods). As part of this step, features are removed that will not provide useful information to build the model. For example, strongly correlated features provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).

|   **TODO: check why ASVs were removed in preprocessing (nzv?)**
|   (depending on answer can discuss why features were removed - e.g. only in one or a few samples)

Mikropml{}

## Discussion/Conclusions (\~250 words)

These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool. In general, we found that finer resolution (e.g. ASVs) does not add additional sensitivity to model prediction and at the ASV level actually impedes model performance due to the sparsity of shared taxa.

-   Utilizing genus or family level data could allow for merging data generated from different 16S regions or sequencing platforms. (is this true?)

While predictive models utilizing microbiome data alone are not that great, microbiome in combination with FIT results are better than either alone {}. This analysis wanted to quantify difference between

@@

To compare performance between taxonomic levels and models, P values were calculated as previously described {begum}. Another output from the mikropml package is the permuted feature importance which is quantified by iteratively permuting each feature in the model and assessing the change in model performance. Features are presumed to be important if the performance of the model, measured by the AUROC, decreases when that feature is permuted. Ranking of feature importance was determined by ordering the features based on the change in AUROC where features with a larger decrease in AUROC are ranked higher in importance.

>>>>

---
bibliography: references.bib
link-citations: true
csl: msystems.csl
output:
  pdf_document:
    keep_tex: true
  word_document: 
    reference_docx: word_formatting.docx
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{multirow}
---

@@

Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States [@siegel2020]. Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e. colonoscopy), many people do not comply with recommended screening guidelines [@garcía2011a] prompting a need for low cost, non-invasive detection methods. A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e. advanced adenomas) and carcinomas. Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}. However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown. Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs [@callahan2017]. However, whether the additional resolution provided by ASVs is useful for ML classification is unclear {}. Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. Topçuoğlu *et al* [@topçuolu2020] recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml). This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.

@@

Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quantified how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or carcinoma) using five machine learning methods with the Mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but borderline equivalent to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A). These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.

While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUC can have different ROC curve shapes [@lobo2008]. Depending on the intended implementation of the model, one may want to optimize the true-positive rate (or sensitivity) over the false-positive rate (or 1-specificity), or vice versa. In this case, the optimal model will detect as many true positives (people with SRNs) as possible. To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%. The highest sensitivity values were observed for Family, Genus, and OTU level data (Figure 1B), consistent with the AUROC results. Phylum, Class, Order, and ASV sensitivity values were all significantly lower than Family, Genus, and OTU sensitivity values (Figure 1B). This analysis further supports that finer resolution is does not improve SRN detection.

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels. The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 3). Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods). As part of this step, features are removed that will not provide useful information to build the model. For example, perfectly correlated features provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1). While the resolution provided by ASVs is useful in certain contexts {}, these results suggest that the resolution is too fine for use in machine learning classification of SRNs based on microbiome composition.

A look into the most important taxa at each level for classifying samples reveals some nesting where several genera and their higher taxonomic classifications are in the top 10 most important taxa (Supplemental Figure 4). For example, the genus *Gemella* is an important taxa at the genus and OTU levels and its higher classifications are also important (*Firmicutes* > *Bacilli* > *Bacillales* > *Bacillales Incertae Sedis XI* > *Gemella*). *Fusobacterium* displays a similar pattern, except that the family level classification (*Fusobacteriaceae*) importance is ranked 16th out of 54 families, a little beyond the top 10. In the case of unclassified *Lachnospiraceae*, there are several OTUs with this label that are in the top 10, however at the Genus level this taxon is ranked lower in importance (21st out of 115 genera) suggesting there may be some benefit to separating different taxonomic groupings within *Lachnospiraceae*.

These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool is warranted. In general, we found that finer taxonomic resolution (e.g. OTU and ASV) does not add additional sensitivity to predicting SRNs based on microbiome composition. Additionally, at the ASV level the fine resolution actually impedes model performance due to the sparsity of shared taxa and leads to decreased model performance. The tendancy for ASV level annotation to split single bacterial genomes into multiple taxa {Pat mSphere 2021} could also be a contributing factor to the sparsity of shared taxa. Overall, either Family, Genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome. A potential benefit of utilizing genus or family level data could be that it may allow for merging data generated from different 16S regions or sequencing platforms.

@@

To compare performance between taxonomic levels and models, P values were calculated as previously described {begum}. To compare sensitivity at 90% specificity, probabilities on the test dataset were collected for each seed and used to construct ROC curves (R pROC::roc). From the ROC curves The sensitivity at a specificity of 90% was pulled for each seed. An optional output from the mikropml package is the permuted feature importance which is quantified by iteratively permuting each feature in the model and assessing the change in model performance. Features are presumed to be important if the performance of the model, measured by the AUROC, decreases when that feature is permuted. Ranking of feature importance was determined by ordering the features based on the change in AUROC where features with a larger decrease in AUROC are ranked higher in importance.

### 2021-08-13 paper/manuscript.tex
first full draft

\hypertarget{introduction-250-words}{%
\subsection{Introduction (\textasciitilde250
words)}\label{introduction-250-words}}

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States
(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
are essential to increase survival rates, but for a variety of reasons
including the invasiveness and high cost of screening
(i.e.~colonoscopy), many people do not comply with recommended screening
guidelines (\protect\hyperlink{ref-garcuxeda2011a}{2}) prompting a need
for low cost, non-invasive detection methods. A growing body of research
points to the gut microbiome as a promising target for non-invasive
screening to detect screen relevant neoplasias (neoplasms?) (SRNs)
consisting of pre-cancerous polyps (i.e.~advanced adenomas) and
carcinomas. Efforts to realize the diagnostic potential of the gut
microbiome in detecting SRNs have focused on machine learning (ML)
methods using abundances from operational taxonomic unit (OTU)
classifications based on amplicon sequencing of the 16S rRNA gene \{\}.
However, whether this is the optimal taxonomic resolution for
classifying SRNs from microbiome data is unknown. Additionally, recent
work has pushed for the use of amplicon sequence variants (ASVs) to
replace OTUs for marker-gene analysis because of the improved resolution
with ASVs (\protect\hyperlink{ref-callahan2017}{3}). However, whether
the additional resolution provided by ASVs is useful for ML
classification is unclear \{pat preprint?\}. Since ML classification
relies on consistent differences between groups, its possible that the
resolution at the ASV level is too individualized to accurately
differentiate groups. Topçuoğlu \emph{et al}
(\protect\hyperlink{ref-topuxe7uolu2020}{4}) recently demonstrated
effective application of machine learning (ML) to microbiome based
classification problems and developed a framework for applying ML
practices in a more reproducible way (mikropml). This analysis utilizes
the reproducible framework developed by Topçuoğlu \emph{et al} to
quantify which ML method and taxonomic level produce the best performing
classifier for detecting SRNs from microbiome data.

\hypertarget{results-700-words}{%
\subsection{Results (\textasciitilde700
words)}\label{results-700-words}}

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated abundance tables
with Mothur \{\} annotated to Phylum, Class, Order, Family, Genus, OTU
and ASV levels. Using the taxonomic abundance data, we quantified how
accurately samples could be classified as ``normal'' or ``SRN''
(i.e.~advanced adenoma or carcinoma) using five machine learning methods
with the Mikropml R package (methods). Across the five machine learning
methods tested, model performance tended to increase with taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Random forest (RF) was
consistently the top performer at most taxonomic levels. Within the RF
model, the highest AUROCs were observed for family (median AUROC:
0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
data with no significant difference between the three (Figure 1).
Performance with ASVs (median AUROC: 0.676) was significantly lower than
OTU (p \textless{} 0.01), but borderline equivalent to family (p = 0.06)
and genus (p = 0.05) levels. These results were consistent for ASVs
generated with DADA2 (\protect\hyperlink{ref-callahan2016}{5}) (median
AUROC: 0.66). These results suggest that finer taxonomic resolution is
not necessarily better for differentiating a persons cancer status from
the microbiome.

While AUROC values between models is a useful way to compare the overall
model performance across all thresholds, they can be misleading
(\protect\hyperlink{ref-lobo2008}{6}). Depending on the intended
implementation of the model, one may want to optimize the true-positive
rate(or sensitivity) over the false-postivite rate (or specificity). To
further compare the model performance across taxonomic levels we
compared the sensitivity of the models

\begin{itemize}
\tightlist
\item
  on average, how many positives are missed with ASV?
\end{itemize}

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus/OTU data there is enough data and variation but at the
ASV level, the data is too specific to individuals and doesn't overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa are present in greater than 75\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite is observed at the OTU and ASV level where 60\% and 53\% of
taxa respectively are only present in less than 25\% of samples
(Supplemental Figure 3). While the resolution provided by ASVs is useful
in

~~\textbf{TODO: what are important taxa, patterns along levels (e.g fuso
always in top )}

Of note, the ML pipeline includes a pre-processing step that occurs
prior to training and classifying the ML models (methods). As part of
this step, features are removed that will not provide useful information
to build the model. For example, strongly correlated features provide
the same information to build the model and thus can be collapsed.
Additionally, features with zero or near-zero variance will not help the
model differentiate groups and thus can be removed. Interestingly,
despite starting with 104106 features at the ASV level, only 478 (0.5\%)
remained after pre-processing. At the OTU level, 20079 of the 705
features (3.5\%) remained after preprocessing (Table 1).

~~\textbf{TODO: check why ASVs were removed in preprocessing (nzv?)}\\
\hspace*{0.333em}\hspace*{0.333em}(depending on answer can discuss why
features were removed - e.g.~only in one or a few samples)

Mikropml\{\}

\hypertarget{discussionconclusions-250-words}{%
\subsection{Discussion/Conclusions (\textasciitilde250
words)}\label{discussionconclusions-250-words}}

These results demonstrate that consideration of the appropriate
taxonomic resolution for utilizing the microbiome as a predictive tool.
In general, we found that finer resolution (e.g.~ASVs) does not add
additional sensitivity to model prediction and at the ASV level actually
impedes model performance due to the sparsity of shared taxa.

\begin{itemize}
\tightlist
\item
  Utilizing genus or family level data could allow for merging data
  generated from different 16S regions or sequencing platforms. (is this
  true?)
\end{itemize}

While predictive models utilizing microbiome data alone are not that
great, microbiome in combination with FIT results are better than either
alone \{\}. This analysis wanted to quantify difference between

@@

To compare performance between taxonomic levels and models, P values
were calculated as previously described \{begum\}. Another output from
the mikropml package is the permuted feature importance which is
quantified by iteratively permuting each feature in the model and
assessing the change in model performance. Features are presumed to be
important if the performance of the model, measured by the AUROC,
decreases when that feature is permuted. Ranking of feature importance
was determined by ordering the features based on the change in AUROC
where features with a larger decrease in AUROC are ranked higher in
importance.

>>>>

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States
(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
are essential to increase survival rates, but for a variety of reasons
including the invasiveness and high cost of screening
(i.e.~colonoscopy), many people do not comply with recommended screening
guidelines (\protect\hyperlink{ref-garcuxeda2011a}{2}) prompting a need
for low cost, non-invasive detection methods. A growing body of research
points to the gut microbiome as a promising target for non-invasive
screening to detect screen relevant neoplasias (neoplasms?) (SRNs)
consisting of pre-cancerous polyps (i.e.~advanced adenomas) and
carcinomas. Efforts to realize the diagnostic potential of the gut
microbiome in detecting SRNs have focused on machine learning (ML)
methods using abundances from operational taxonomic unit (OTU)
classifications based on amplicon sequencing of the 16S rRNA gene \{\}.
However, whether this is the optimal taxonomic resolution for
classifying SRNs from microbiome data is unknown. Additionally, recent
work has pushed for the use of amplicon sequence variants (ASVs) to
replace OTUs for marker-gene analysis because of the improved resolution
with ASVs (\protect\hyperlink{ref-callahan2017}{3}). However, whether
the additional resolution provided by ASVs is useful for ML
classification is unclear \{\}. Since ML classification relies on
consistent differences between groups, its possible that the resolution
at the ASV level is too individualized to accurately differentiate
groups. Topçuoğlu \emph{et al}
(\protect\hyperlink{ref-topuxe7uolu2020}{4}) recently demonstrated
effective application of machine learning (ML) to microbiome based
classification problems and developed a framework for applying ML
practices in a more reproducible way (mikropml). This analysis utilizes
the reproducible framework developed by Topçuoğlu \emph{et al} to
quantify which ML method and taxonomic level produce the best performing
classifier for detecting SRNs from microbiome data.

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated abundance tables
with Mothur \{\} annotated to Phylum, Class, Order, Family, Genus, OTU
and ASV levels. Using the taxonomic abundance data, we quantified how
accurately samples could be classified as ``normal'' or ``SRN''
(i.e.~advanced adenoma or carcinoma) using five machine learning methods
with the Mikropml R package (methods). Across the five machine learning
methods tested, model performance tended to increase with taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Random forest (RF) was
consistently the top performer at most taxonomic levels. Within the RF
model, the highest AUROCs were observed for family (median AUROC:
0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
data with no significant difference between the three (Figure 1A,
Supplemental Figure 2). Performance with ASVs (median AUROC: 0.676) was
significantly lower than OTUs (p \textless{} 0.01), but borderline
equivalent to family (p = 0.06) and genus (p = 0.05) levels (Figure 1A).
These results were consistent for ASVs generated with DADA2
(\protect\hyperlink{ref-callahan2016}{5}) (median AUROC: 0.66). These
results suggest that finer taxonomic resolution is not necessarily
better for identifying individuals with SRNs based on microbiome
composition.

While comparing AUROC values between models is a useful way to assess
the overall model performance, they summarize the performance across all
thresholds and can be misleading since models with the same AUC can have
different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{6}).
Depending on the intended implementation of the model, one may want to
optimize the true-positive rate (or sensitivity) over the false-positive
rate (or 1-specificity), or vice versa. In this case, the optimal model
will detect as many true positives (people with SRNs) as possible. To
further compare the model performance across taxonomic levels we
compared the sensitivity of the models at a specificity of 90\%. The
highest sensitivity values were observed for Family, Genus, and OTU
level data (Figure 1B), consistent with the AUROC results. Phylum,
Class, Order, and ASV sensitivity values were all significantly lower
than Family, Genus, and OTU sensitivity values (Figure 1B). This
analysis further supports that finer resolution is does not improve SRN
detection.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus/OTU data there is enough data and variation but at the
ASV level, the data is too specific to individuals and doesn't overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa are present in greater than 75\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite is observed at the OTU and ASV level where 60\% and 53\% of
taxa respectively are only present in less than 25\% of samples
(Supplemental Figure 3). Of note, the ML pipeline includes a
pre-processing step that occurs prior to training and classifying the ML
models (methods). As part of this step, features are removed that will
not provide useful information to build the model. For example,
perfectly correlated features provide the same information to build the
model and thus can be collapsed. Additionally, features with zero or
near-zero variance will not help the model differentiate groups and thus
can be removed. Interestingly, despite starting with 104106 features at
the ASV level, only 478 (0.5\%) remained after pre-processing. At the
OTU level, 20079 of the 705 features (3.5\%) remained after
preprocessing (Table 1). While the resolution provided by ASVs is useful
in certain contexts \{\}, these results suggest that the resolution is
too fine for use in machine learning classification of SRNs based on
microbiome composition.

A look into the most important taxa at each level for classifying
samples reveals some nesting where several genera and their higher
taxonomic classifications are in the top 10 most important taxa
(Supplemental Figure 4). For example, the genus \emph{Gemella} is an
important taxa at the genus and OTU levels and its higher
classifications are also important (\emph{Firmicutes} \textgreater{}
\emph{Bacilli} \textgreater{} \emph{Bacillales} \textgreater{}
\emph{Bacillales Incertae Sedis XI} \textgreater{} \emph{Gemella}).
\emph{Fusobacterium} displays a similar pattern, except that the family
level classification (\emph{Fusobacteriaceae}) importance is ranked 16th
out of 54 families, a little beyond the top 10. In the case of
unclassified \emph{Lachnospiraceae}, there are several OTUs with this
label that are in the top 10, however at the Genus level this taxon is
ranked lower in importance (21st out of 115 genera) suggesting there may
be some benefit to separating different taxonomic groupings within
\emph{Lachnospiraceae}.

These results demonstrate that consideration of the appropriate
taxonomic resolution for utilizing the microbiome as a predictive tool
is warranted. In general, we found that finer taxonomic resolution
(e.g.~OTU and ASV) does not add additional sensitivity to predicting
SRNs based on microbiome composition. Additionally, at the ASV level the
fine resolution actually impedes model performance due to the sparsity
of shared taxa and leads to decreased model performance. The tendancy
for ASV level annotation to split single bacterial genomes into multiple
taxa \{Pat mSphere 2021\} could also be a contributing factor to the
sparsity of shared taxa. Overall, either Family, Genus, or OTU level
taxonomy appear to perform equally for predicting subjects with SRNs
based on the composition of the gut microbiome. A potential benefit of
utilizing genus or family level data could be that it may allow for
merging data generated from different 16S regions or sequencing
platforms.

@@

To compare performance between taxonomic levels and models, P values
were calculated as previously described \{begum\}. To compare
sensitivity at 90\% specificity, probabilities on the test dataset were
collected for each seed and used to construct ROC curves (R pROC::roc).
From the ROC curves The sensitivity at a specificity of 90\% was pulled
for each seed. An optional output from the mikropml package is the
permuted feature importance which is quantified by iteratively permuting
each feature in the model and assessing the change in model performance.
Features are presumed to be important if the performance of the model,
measured by the AUROC, decreases when that feature is permuted. Ranking
of feature importance was determined by ordering the features based on
the change in AUROC where features with a larger decrease in AUROC are
ranked higher in importance.
